---
title: "Chapter 5"
author: "Agata Dominowska"
date: "6 December 2017"
output: html_document
---
# Dimensionality reduction techniques
*The objective of this week was to learn the basics of two data science based ways of reducing the dimensions, PCA and MCA. *

## 1. Reading and exploring the data
*Task 1 and 2*
The original data, coming from the United Nations Development Programme, is composed of 2 data sets measuring human development and gender inequality for a set of countries (to be more precise, [Human Development Index (HDI)]( http://hdr.undp.org/en/composite/HDI) and [Gender Inequality Index (GII)](http://hdr.undp.org/en/composite/GII)). It is quite recent, as most of it comes from year 2015. 
The data used in this part has been modified on the basis of the original data, however.

Under these links my [processed data in the .csv format](https://github.com/aadomino/IODS-project/blob/master/data/human.csv) and the [script used to process the data](https://github.com/aadomino/IODS-project/blob/master/data/create_human.R) can be found.

```{r}
human <- read.csv(file = "C:\\Users\\P8Z77-V\\Documents\\GitHub\\IODS-project\\data\\human.csv", sep = ",", header = TRUE)
str(human)
dim(human)
```

The data includes 8 variables (*Edu2.FM*, *Labo.FM*, *Life.Exp*, *Edu.Exp*, *GNI*, *Mat.Mor*, *Ado.Birth*, *Parli.F*) of human development and inequality for the 155 countries (*Country* column) that had the complete data. Each row is represented by data from a different country. These variables tell about the following phenomena, respectively:

* The percentage of women with at least secondary education 
* The percentage of women participating in labour force 
* Life expectance at birth 
* Expected years of schooling 
* GNI (gross national income) per capita 
* Maternal mortality ratio 
* Adolescent birth rate 
* Female share of parliamentary seats 

The summary of the data shows the distributions and relations of the observations - we can observe e.g. the minimum, maximum, median and mean values of the variables:

```{r}
summary(human)
```

The striking impression about this overview is that there is a lot of variation between the countries, which is shown by the min and max values for the variables in the data. For example, the presence of women in the parilament ranges from 0 to 57.5%, the maternal mortality from 1 to 1100 (per 100 000 births), lthe proportion of women with a least secondary education ranges from 0.9% to 100%. 

The more detailed look at the data with ggpairs function:

```{r}
# Access GGally
library(GGally)

# Visualize the 'human' variables
ggpairs(human)
```

And a clearer look into the most prominent correlations with the correlation matrix visualisation (correlation plot):

```{r}
# Calculate the correlation matrix and round it to include just 2 digits
cor_matrix<-cor(human) %>% round(digits=2)
 
# Visualize the correlation matrix with a correlations plot
corrplot(cor_matrix, method="circle", type = "upper", cl.pos = "b", tl.pos = "d", tl.cex = 0.6)
```

Some of the varibles have quite strong correlations with each other. For example, secondary education correlates positively with life expectancy and (quite understandably) with more years of schooling; it is however negatively correlated with adolescent births and maternal deaths. Life expectancy and expected length of education also correlate positively between themselves, and so do maternal mortality with adolescent births; these also are connected with, respectively, the high or low standard of living in a given country - GDI is correlated  slightly negatively with both maternal deaths and adolescent births. It seems that a bad situation of women in a country correlates with other negative issues.


##2. Principal component analysis (PCA)
*Tasks 3, 4 and 5*

The next step is to perform principal component analysis (PCA) using the singular value decomposition (SVD) on the not standardized human data and show the variability captured by the principal components. Then, we will draw a biplot displaying the observations by the first two principal components (PC1 coordinate in x-axis, PC2 coordinate in y-axis), along with arrows representing the original variables.  

```{r}
pca_human <- prcomp(human)
summary(pca_human)

biplot(pca_human, choices = 1:2, cex = c(0.5, 0.8), col = c("grey40", "deeppink2"), main = "Biplot, unscaled human data")
```

It is clear that standard deviations (or variablity) of the principal components are very different in magnitude. From the summary of the PCA results and the variance captured by each component it is clear that all of the variability in the original features is captured by the first principal component.PCA is a method sensitive to the relative scaling of the original features and takes features with larger variance to be more important than features with smaller variance.

The biplot is not easy to read. Most data are at the top right corner of it. Since the variables have not been standardized, gross national income (GNI) is the only significant variable because of its large variance. No variablility is captyred by PC2 or others, so other relationships are not really shown and hence this plot is not very informative.

Let's standardize the variables in the human data and repeat the above analysis and then see what happens.

```{r}
# Scale
human_std <- scale(human)

# Estimation
pca_human <- prcomp(human_std)
summary(pca_human)

# rounded percetanges of variance captured by each PC
pca_pr <- round(100*pca_human$importance[2, ], digits = 1)

# create object pc_lab to be used as axis labels
pc_lab <- paste0(c("Underdevelopment", "Gender equality"), " (", pca_pr, "%)")

biplot(pca_human, choices = 1:2, cex = c(0.5, 0.8), col = c("grey40", "deeppink2"), xlab = pc_lab[1], ylab = pc_lab[2], main = "Biplot, scaled human data")
```

After scaling the variables, it becomes visible that the values of each variable are more balanced and sort of flattened, and the second biplot is much easier to interpret: the viariability of each principal component is more evenly distrubited, yet the prominence of PC1 is still visible.

The biplot could be interpreted as follows: the PC1 compoment reflects low human development with variables of low life expectancy, low education, high maternal mortality. The praliamentary participation, labour market participation and GNI are not correlated with the PC1, but are with PC2, the second component which seems to reflect gender equality. The second principal component is relatively less important (see the length of the arrows.)

##3. Multiple correspondence analysis on FactoMineR
*Task 6*

Here, we will load the tea dataset from the package Factominer and first explore the data briefly, look at the structure and the dimensions of the data and visualize it. Then we will do Multiple Correspondence Analysis on the tea data. 

```{r}
library(FactoMineR)
# load data
data("tea")

str(tea)
dim(tea)
```

This tea consumption dataset has 300 observations and 36 variables, all of which are factor or categorical variables. The variables include questions on background factors (sex, age), habits related to drinking tea (frequency, sugar, at work etc), and attitudes towards this beverage (friendliness, exciting, spirituality...).

6 variables are selected for the MCA (following DataCamp):

```{r}
# column names to keep in the dataset
keep_columns <- c("Tea", "How", "how", "sugar", "where", "lunch")

# select the 'keep_columns' to create a new dataset
tea_time <- dplyr::select(tea, one_of(keep_columns))
```

The MCA of the tea dataset:

```{r}
# multiple correspondence analysis
mca <- MCA(tea_time)

# summary of the model
summary(mca)

# visualize MCA
plot(mca, invisible=c("ind"), habillage = "quali")
```
MCA can be used to detect patterns and tendencies in qualitative data (which we humanists love so much). 

The MCA plots group the observations. The variables how and where (the purchase place and the packaging form) are close to each other and are more prominent than others. 

The last plot shows how the variables relate to the dimensions. It can be observed that the majority of individuals are found in the middle of the plot and there are no outlier cases. Buying tea from tea shops and chain stores, buying tea bags from chain stores and buying unpacked tea from tea shops are pairs that kind of go together - these are the most popular purchase patterns. Earl Grey tea is close to milk, while green tea definitely is not.


